\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Early and correct diagnosis of patient symptoms is critical to enhance access to health services and lessen unnecessary visits to the doctor. This thesis offers a controlled comparison of five machine learning classifiers --- Naive Bayes, Logistic Regression, Support Vector Machine, Random Forest, and XGBoost --- on the problem of diagnosing diseases based on binary symptom vectors. This research seeks to address two questions: which classifier is most accurate at predicting diseases based on their symptoms, and does the prediction change as the dataset increases in size and imbalance.

Two publicly available Kaggle datasets are used. Dataset 1 is a small, edited set of around 5,000 records of 132 symptoms and 41 diseases. Dataset 2 is much larger, having around 246,000 records and 727 disease classes, which represents a more realistic amount of class symptom overlap and imbalance. In Dataset 1, grid search and stratified cross-validation are used to tune the hyperparameters. On Dataset 2, fixed hyperparameters are employed to make the experiments computationally feasible. All models are evaluated using accuracy, precision, recall, and macro-averaged F1-score, with class weighting used instead of synthetic oversampling since interpolation-based oversampling strategies, like SMOTE, are ill-suited to binary feature spaces.

The findings indicate that there is no best classifier that can be used in both environments. On the smaller data, SVM and Logistic Regression achieve perfect scores, with Naive Bayes and Random Forest being very competitive at about 98\% accuracy. On the bigger dataset, however, performance rankings shift significantly: Naive Bayes has the highest accuracy (84.11\%), and Logistic Regression and XGBoost are still competitive, whereas Random Forest falls to a very low value, implying that the choice of a model should be based on the characteristics of the actual data and not a general hypothesis that more complex methods dominate. Error analysis also reveals that the majority of misclassifications are between diseases with similar profiles of symptoms; this is medically explainable and arguably acceptable in a triage-style booking situation where the same conditions are treated by the same specialist.

In addition to crude metric comparisons, this thesis discusses the real-world trade-offs that influence actual deployment: interpretability, computational cost, and adherence to regulatory standards, including the EU AI Act and the GDPR. Based on these findings, Logistic Regression is suggested where interpretability is important, Naive Bayes as an efficient, powerful baseline on large, imbalanced data, and ensemble techniques only when they justify their added cost by empirical gains. The study also suggests that the deployment of any such system should only be possible under external clinical validation, calibrated confidence thresholds, and human oversight. The main limitations are the absence of external clinical validation, the binary representation of symptoms, and the reduction of the prediction task to a single-disease formulation, all of which establish definite directions for future work.

\vspace{0.5cm}
\noindent\textbf{Keywords:} machine learning, disease prediction, symptom-based classification, healthcare, Random Forest, XGBoost, comparative study
