\chapter{Introduction}
\label{ch:introduction}

\section{Background and Motivation}
\label{sec:background}

It is a universal issue in healthcare systems worldwide because patients are usually not aware of the doctor they need to see. A patient who has had a lingering cough may see a general practitioner and be referred to a pulmonologist, and only after several weeks attending to the referral appointment does the patient get a diagnosis. This lack of efficiency is time and resource wasting on both ends. Patients end up waiting to get proper care and health care providers use up appointment time in cases that would have been diverted in a more efficient manner. In the digital health strategy presented by the World Health Organization, primary healthcare systems across the globe struggle to match patients with the right specialist care at the right time, especially in underserved areas where providers are scarce to provide care to patients on time and in the most suitable way possible.\citep{WHO2021}.

It is not only an issue of waiting times. Less efficient routes also result in poor health outcomes. A patient with a severe case may take weeks to visit a GP, and after another week, he or she may have to wait again before he or she receives a specialist referral, which postpones diagnosis and treatment. This delay in certain instances can spell the difference between a successful treatment and irreparable damage.

General practitioners on the other side waste a lot of time on cases which might be dealt with directly by specialists. This introduces a bottleneck in the healthcare system and makes the doctors spend less time on patients who actually require general care. Everyone would benefit with a more efficient system of triage: patients can see the appropriate specialist quicker, GPs get additional capacity to help more patients requiring their professional help, and specialists may also see the people who actually need their services.

Machine learning is a possible solution. With the help of training classifiers on the datasets of symptoms and their related diagnoses, one can create models that could predict the probable diseases given a collection of reported symptoms \citep{Rajpurkar2022}. A healthcare booking system could include such a model. As soon as a patient makes an appointment request, the system requests the patient to complete a questionnaire of symptoms, executes the classifier using their answers, and proposes the most suitable specialist depending on the predicted disease. E.g. a patient with some type of skin related symptoms may be instructed to make a direct appointment with a dermatologist instead of visiting a general practitioner.

There are a number of benefits of this type of system. It is 24/7 without the need of a nurse or a doctor manually triaging each request. It is scaled to serve a large number of patients at any given time. It may be implemented in a web or mobile application and thus can be available in remote locations where face-to-face triage is not a possibility. And when done properly it may help to decrease waiting times and increase patient outcomes.

Nevertheless, the construction of such a system poses key questions. What machine learning classifier is applicable? What level of accuracy must the model have before it can be safely deployed? What does the model do when it is not certain? What do you do with infrequent diseases that are not represented in the training data? These are the questions this thesis tries to address.

\subsection{Machine Learning for Disease Prediction}

Symptom-based disease prediction has been conducted with several classifiers, including basic such as Naive Bayes, as well as the ensemble models such as Random Forest and XGBoost \citep{Uddin2019}. These approaches have various advantages and disadvantages. Naive Bayes is quick and simple to understand but contains strong assumptions of feature independence which may not be true of symptom data. Also interpretable and effective with problems which are linearly separable, Logistic Regression cannot learn complex interactions between symptoms without manual feature engineering. The Support Vector Machines are able to cope with non-linear decision boundaries but they are slow to train when dealing with large datasets. Random Forest and XGBoost are the robust ensemble techniques that can be best performing, whereas they are more complex and cannot be easily interpreted.

As I was going through the literature, I was surprised that most of the studies only test one or two classifiers at the same time and it is difficult to determine which method really works best. Various papers have varied datasets, varied preprocessing methods, and varied evaluation measures thus making it hard to compare outcomes of various papers. Other studies achieve such high numbers of accuracy without stating how they dealt with class imbalance, which is a significant concern in medical datasets, where certain diseases are far less prevalent than others.

Minimal work is also done on the behavior of these classifiers when the size of the dataset varies in a significant manner. There are classifiers that have been known to perform well on small data sets, but level off as more data is available, whereas others are still improving. This scaling behavior is significant to understand since the amount of training data will be vastly different in various deployment situation.

\subsection{Why This Work Matters}

This thesis provides a controlled comparison of five commonly used machine learning classifiers on the same symptom-disease prediction task, using the same preprocessing and evaluation procedures. Using two very different-size datasets, I can observe the performance with respect to the availability of data. With the help of various evaluation metrics (accuracy, precision, recall, F1-score), I will be able to measure various attributes of the classifier performance instead of a single number.

In addition to the technical comparison, this thesis explains the practical and ethical factors that would be of concern in a real deployment. Which one is better; the more accurate but less interpretable model or the less accurate but transparent one? What are the regulatory provisions in the EU AI Act and GDPR? What do you do with the threat of making a false prediction and hurting a patient? These are not issues that can be resolved by machine learning, and need to be contributed to by doctors, ethicists, regulators, and even patients themselves. This thesis is not able to address all of them but attempts to set them into perspective so that they can be addressed in the future.

\section{Problem Statement}
\label{sec:problem_statement}

No systematic comparison of several machine learning classifiers to predict diseases based on symptoms under controlled conditions is done. The literature, as found through the systematic review of the literature by \citet, identifies a single algorithm, or small group of models (usually two or three models) on various datasets with various preprocessing steps, which complicated the comparison of results across studies. Whenever one study claims that Random Forest gets 92-percent accuracy, and another study claims that XGBoost gets 94-percent accuracy, you cannot say that XGBoost is superior because they may have used different data sets, different train-test splits, different hyperparameter tuning processes, or different evaluation metrics.

Class imbalance - the fact that certain diseases are much more frequent than others in the data - is not treated consistently across the literature \citep{Haixiang2017,Thabtah2020}. Methods include oversampling methods such as SMOTE \citep{Chawla2002} and class weighting methods, as well as methods which do not make use of an explicit balancing method. It is difficult to tell with this, whether the reported performance numbers represent real classifier quality or simply good performance on common diseases and ignoring rare ones.

The other gap is the absent multi-scale evaluation. The majority of studies are tested on one dataset, thus we are not aware of the ability to apply the results to other data sizes. What works well on 5,000 examples may not generalize to 246,000 examples, or even do much better. This scaling behavior is of importance in terms of practical deployment.

Lastly, the vast majority of research results are in terms of predictive performance (accuracy, F1-score) and do not address the more practical trade-offs that are important to actual deployment: training time, interpretability, regulatory compliance, behavior with rare diseases, and what the model does when it is uncertain. These considerations may be equally relevant to the determination of which classifier to use as raw accuracy.

\section{Research Question}
\label{sec:research_question}

This thesis addresses the following research question:

\begin{quote}
\textit{What is the comparison of Naive Bayes, Logistic Regression, Support Vector Machine, Random Forest, and XGBoost in regards to accuracy, precision, recall and F1-score when used in symptom-based disease prediction and what is the best trade-off between predictive performance and applicability in healthcare decision support?}
\end{quote}

\section{Objectives}
\label{sec:objectives}

To answer this question, I pursue the following objectives:

\begin{enumerate}
    \item Review the existing literature on machine learning for symptom-based disease prediction.
    \item Describe and preprocess two publicly available Kaggle datasets for symptom-based disease classification.
    \item Implement and train five classifiers --- Naive Bayes, Logistic Regression, Support Vector Machine, Random Forest, and XGBoost --- under identical experimental conditions.
    \item Evaluate and compare these classifiers using accuracy, precision, recall, and F1-score, and analyze error patterns using confusion matrices and feature importance.
    \item Discuss what the results mean for healthcare booking and triage.
\end{enumerate}

\section{Scope and Limitations}
\label{sec:scope}

This thesis focuses on predicting disease categories from binary symptom vectors using supervised classification. The scope includes:

\begin{itemize}
    \item Comparing five machine learning classifiers (Naive Bayes, Logistic Regression, SVM, Random Forest, XGBoost) under identical experimental conditions
    \item Testing on two publicly available Kaggle datasets of different sizes
    \item Handling class imbalance through class weighting
    \item Evaluating performance using multiple metrics (accuracy, precision, recall, F1-score)
    \item Analyzing error patterns through confusion matrices and feature importance
    \item Discussing practical deployment considerations including interpretability, regulatory requirements, and ethical concerns
\end{itemize}

The thesis does \textbf{not} cover:

\begin{itemize}
    \item Deep learning methods (neural networks, transformers) --- I focus on traditional ML classifiers that are more interpretable and require less training data
    \item The disease-to-specialist mapping layer --- I assume this mapping exists but don't implement or evaluate it
    \item Treatment recommendation --- I only predict disease categories, not what treatment the patient should receive
    \item Real-time clinical deployment --- the evaluation is offline using historical datasets, not tested with real patients
    \item Multi-label classification --- I assume each patient has exactly one disease, though in reality patients can have multiple conditions
    \item Symptom severity or temporal information --- the datasets only provide binary symptom presence/absence
\end{itemize}

The analysis is based on two publicly available datasets on Kaggle and is not clinical and does not use real patient data on hospitals or clinics. The performance figures found in this thesis will need to be taken as upper limits. With realistic data of patients, performance would probably be poorer because the symptom reporting would be more noisy, the disease distribution would not be the same, and other variables that are not represented in the cleaned Kaggle datasets would occur.

I speak about the implication in terms of how the healthcare booking works at a conceptual level and not in terms of a deployed system. The development of a full-fledged booking system would imply the integration with appointment scheduling software, user interface design, regulatory approval, and collaboration with healthcare providers all of which are not within the scope of a bachelor thesis.

\section{Thesis Structure}
\label{sec:thesis_structure}

Chapter~\ref{ch:literature_review} reviews the relevant literature. Chapter~\ref{ch:data_description} describes the two datasets and the preprocessing pipeline. Chapter~\ref{ch:methodology} presents the five classifiers, the handling of class imbalance, and the evaluation metrics. Chapter~\ref{ch:results} reports the experimental results. Chapter~\ref{ch:discussion} discusses the findings, limitations, and ethical considerations. Chapter~\ref{ch:conclusion} summarizes the contributions and outlines future work.

\vspace{1cm}

\noindent\textbf{Thesis Statement:} This thesis compares five machine learning classifiers for symptom-based disease prediction across two datasets of different scales, and argues that the best model for real-world deployment depends not just on accuracy but also on interpretability, computational cost, and how the ranking changes when the dataset becomes larger and more imbalanced.
