\chapter{Data Description}
\label{ch:data_description}

In this chapter, the author tells us about the two publicly available datasets considered in this thesis, their nature, and the preprocessing that was carried out on them to classify them. Both of them are also based on Kaggle and are symptom-based disease prediction tasks on various scales.

\section{dataset 1: Disease Symptom Prediction Dataset.}
\label{sec:dataset1}

The first data, the Disease Symptom Prediction Dataset \citep{Kaggle_Disease2020} has about 5,000 records of patients. A combination of symptoms is recorded on each record and assigned to one of 41 disease categories. The dataset consists of 132 distinct symptoms in form of columns with 1 representing the existence of a symptom and 0 representing the absence.

The main features of this data are summarized in Table~\ref{tab:dataset1_overview}.

\begin{table}[H]
\centering
\caption{Overview of Dataset~1: Disease Symptom Prediction Dataset}
\label{tab:dataset1_overview}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property}         & \textbf{Value} \\
\midrule
Source                     & Kaggle (itachi9604) \\
Number of records          & $\sim$5,000 \\
Number of features         & 132 (binary symptoms) \\
Number of disease classes  & 41 \\
Feature type               & Binary (0/1) \\
Target variable            & Disease name (categorical) \\
Missing values             & Present in raw data \\
\bottomrule
\end{tabular}
\end{table}

The information was first collected based on several medical resources and also includes additional data such as descriptions of diseases, rating of symptoms by severity, and recommended precautions. I have only applied the symptom-disease mapping, as my main goal is to assess the performance of the classifiers with regard to the core prediction task.

\subsubsection{Class Distribution}

Dataset 1 has a moderately imbalanced class distribution. The majority of disease categories contain 100 to 150 records, with some of the rare diseases having less than 50 examples. The pattern is clear: the top 10 most frequent diseases cover approximately 40 percent of all the records, while the bottom 10 diseases cover less than 10 percent.

Such an imbalance is realistic - it takes into account the reality that certain diseases are literally more prevalent than others. Nevertheless, it will imply that classifiers that have been trained without imbalanced handling will be more likely to predict common diseases and overlook rare ones. I solved this by class weighting (Section~\ref{sec:class_imbalance}).

\subsubsection {Symptom Overlap Among Diseases.}

A notable feature of this data is that a lot of diseases have similar symptoms. As an example, fever is seen in approximately 35 percent of all documentation of various disease categories. This is technically true - fever is an unspecific symptom which may indicate a wide variety of conditions such as the flu, tuberculosis and so on. The problem faced by classifiers is to exploit the combination of symptoms instead of individual symptoms.

I have computed the symptom co-occurrence patterns and discovered that some combinations of the symptoms are very predictive. An example of this is that [yellowish\_skin, loss\_of\_appetite, abdominal\_pain] is strongly associated with liver diseases, while [skin\_rash, itching, nodal\_skin\_eruptions] is strongly linked to dermatological disorders. Random Forest and XGBoost are tree-based classifiers that are effective in finding these patterns since they are able to divide based on several features.

\section{Dataset~2: Diseases and Symptoms Dataset}
\label{sec:dataset2}

The second dataset is the Diseases and Symptoms Dataset \citep{Kaggle_Diseases2023}: it is also much larger, consisting of about 246,000 records of patients. It encompasses more conditions than Dataset~1.

Table~\ref{tab:dataset2_overview} summarizes its key characteristics.

\begin{table}[H]
\centering
\caption{Overview of Dataset~2: Diseases and Symptoms Dataset}
\label{tab:dataset2_overview}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property}         & \textbf{Value} \\
\midrule
Source                     & Kaggle (dhivyeshrk) \\
Number of records          & $\sim$246,000 \\
Number of features         & 377 binary symptoms \\
Number of disease classes  & 727 \\
Feature type               & Binary (0/1) \\
Target variable            & Disease name (categorical) \\
Missing values             & Minimal \\
\bottomrule
\end{tabular}
\end{table}

The increased size of Dataset~2 allows me to visualize how classifiers increase in size with increased data. Almost 50 times bigger than Dataset~1, this dataset has more consistent estimates of performance and less chance of overfitting. It also minimizes the variance of performance estimates between x-validation that increases the reliability of the comparison.

\subsubsection{Class Distribution}

The imbalance pattern of Dataset 2 is also similar to that of Dataset 1, except that due to the larger sample size, even the rarest diseases have hundreds of examples. This is significant since there are certain classifiers (especially Naive Bayes) which require a certain number of examples of each class to make accurate predictions in terms of probabilities.

Funnal infections, common colds, and reactions to drugs are the most prevalent diseases in Dataset~2. The rarer diseases are more specific such as psoriasis and hyperthyroidism. The ratio of class imbalance (most common class/ least common class) is about 50: 1 which is very large but not as severe as some medical datasets I observed in the literature where the ratio can be thousands: one.

\subsubsection{Data Quality}

Data quality is one of the concerns that large web-sourced datasets have. I sampled 100 records of Dataset 2 and ensured that the symptom-disease mappings were sensible. The majority of them were sensible, but there were some instances when the set of symptoms appeared to be incomplete. Indicatively, there was only one record that had two symptoms to be diagnosed with diabetes, yet diabetes is supposed to have more.

Although these quality problems are present on occasion, the size of the dataset is such that some few noise records cannot have much impact on the performance of the classifier. The benefit of 246, 000 examples is greater than the drawback of slightly more noisy labels than a smaller, more well curated dataset.

\section{Dataset Comparison}
\label{sec:dataset_comparison}

The two datasets differ in several important ways beyond size. Table~\ref{tab:dataset_comparison} summarizes the key differences.

\begin{table}[H]
\centering
\caption{Comparison of Dataset~1 and Dataset~2}
\label{tab:dataset_comparison}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Characteristic} & \textbf{Dataset~1} & \textbf{Dataset~2} \\
\midrule
Number of records & $\sim$5,000 & $\sim$246,000 \\
Number of classes & 41 & 727 \\
Number of features & 132 & 377 \\
Class imbalance ratio & $\sim$10:1 & $\sim$50:1 \\
Missing values & Moderate & Minimal \\
Data quality & High (curated) & Good (web-sourced) \\
Smallest class size & $\sim$50 examples & $\sim$500 examples \\
\bottomrule
\end{tabular}
\end{table}

Dataset 1 is quite clean, but small, and therefore, overfitting is a concern, particularly with complex models with lots of parameters. As an example, XGBoost using hundreds of trees can memorize training examples instead of general patterns. Dataset 2 contains additional information but can be more noisy because of the way it was collected. Classifiers should be larger, which should result in them being more robust and less varying in performance estimates.

The reason why I decided to work with two datasets of varying scale is that, I can compare classifiers not only by their absolute performance but also by the way they scale with the availability of data. There are classifiers that are known to perform dismally on small datasets but improve with the availability of more data. The Naive Bayes classifier is frequently used as a baseline on small data sets since it requires relatively few examples to estimate the probability of a particular class. Its performance however levels off as you increase the amount of data, due to the independence assumption that restricts the amount of learning that can be done with the model. On the other hand, ensemble algorithms such as XGBoost can be constantly improved with additional data since they are capable of creating more complicated decision boundaries.

This is important to practical implementation - training data will be extremely different across various healthcare settings. A major hospital could have tens of thousands of historical patient records, whereas a small clinic could only have a few hundred. When a classifier is only effective on large datasets, it will not be effective in low-resource conditions.

The other purpose of using two datasets is to test whether the rankings of the classifiers are stable. When Random Forest outperforms SVM on Dataset~1 and SVM outperforms Random Forest on Dataset~2, this indicates that the ranking has been dataset-dependent and not representative of some overall performance of one approach over the other. Conversely, when Random Forest outperforms SVM on both datasets, then that is more evidence that the former really is superior in this task.

\section{Preprocessing Pipeline}
\label{sec:preprocessing}

I applied the same preprocessing pipeline to both datasets to ensure fair comparison across classifiers. The steps were:

\textbf{Step 1: Data Loading and Inspection.} I loaded each dataset from CSV using \texttt{pandas.read\_csv()} and performed initial inspection. This included checking for structural issues (e.g., mismatched column counts), column types (to make sure symptoms were numeric), missing values, and duplicate records. Duplicate records could artificially inflate performance if the same record appears in both training and test sets, so I removed them before splitting.

\textbf{Step 2: Missing Value Handling.} Missing values in the symptom columns in Dataset 1, which I replaced with 0 (symptom absent), were necessary to match the binary encoding. The assumption is that a symptom that is not stated explicitly as present is absent. This is a sensible default of symptom data. Dataset 2 contained less missing values (less than half a percent of all cells), which I treated in the same manner. The other option would have been to eliminate records that contain missing values which would have eliminated good data.

\textbf{Step 3: Feature Encoding.} Symptoms had to be coded in the form of binary features. Dataset 1 was already in this format where there must have been 0 or 1 in each symptom column. Dataset 2 is also saved in binary symptom table format, so I just load all feature columns except disease label, replace missing values with 0 and retain the dataset-specific symptom names. I do not subject the two datasets to the same feature space, as they are compared independently.

\textbf{Step 4: Label Encoding.} I encoded disease names as integer labels using scikit-learn's \texttt{LabelEncoder}. This maps each disease name to a unique integer (e.g., "diabetes" $\rightarrow$ 0, "pneumonia" $\rightarrow$ 1, etc.). I saved the mapping so I could interpret results later by converting integer predictions back to disease names. All classifiers work with integer labels internally, so this step is necessary.

\textbf{Step 5: Train-Test Split.} I split each dataset into 80\% training and 20\% testing using \texttt{train\_test\_split} with \texttt{stratify=y}. The stratify parameter ensures that the class distribution in the training set matches the class distribution in the test set. For example, if diabetes accounts for 5\% of the full dataset, it will account for 5\% of the training set and 5\% of the test set. This is important for imbalanced datasets because a random split might accidentally put all examples of a rare disease into the training set, leaving none for testing.

I set a fixed random seed (\texttt{random\_state=42}) so that the splits are reproducible. This means anyone can re-run the experiments and get the same train-test split.

\textbf{Step 6: Feature Scaling.} In the case of Logistic Regression and SVM, I used the \texttt{StandardScaler} method to standardize features to mean 0 and unit variance. These classifiers require this since they are scale sensitive to features. And without scaling, the features with larger ranges (not relevant here since all features are binary but would be relevant to features that are continuous) would dominate the model.

In binary features, standardization is not of great importance since all values are either 0 or 1, but I added it as a consistency with the best practices. Naive Bayes and tree-based classifiers (Random Forest, XGBoost) do not require scaling since they are scale-invariant, and thus I used the original binary features to train them.

One notable point: I put the scaler on the training set only, and then used it on both the training set and the test set. Had I scaled the full dataset before splitting, that would be information leaking in the test set to the training process and would provide optimistic performance estimates.

\textbf{Step 7: Cross-Validation Setup.} Cross-Validation Setup. I applied hyperparameter tuning of stratified 5-fold cross-validation on Dataset 1. Stratified splits are important in imbalanced datasets since they maintain the same proportions of classes in each fold. Dataset 2 The hyperparameters are not tuned on a complete grid search, but fixed, since exhaustive tuning would be prohibitively expensive at that scale. Evaluation on both datasets is still done on a held-out test set.

A typical number of folds is 5, which is a trade off between bias and variance on the performance estimate. Reducing the number of folds (e.g., 3) would be much faster but provide a more noisy estimate. A larger number of folds (e.g. 10) would be more precise but more time-consuming to calculate. In the larger Dataset 2, 5-fold cross-validation would already consume a lot of compute time and I did not exceed it.

The complete preprocessing pipeline, including software versions (Table~\ref{tab:software}) and hyperparameter search spaces (Table~\ref{tab:hyperparameters}), is documented in Appendix~\ref{app:supplementary}.
