\chapter{Literature Review}
\label{ch:literature_review}

The chapter discusses the available literature on machine learning to predict diseases. I begin with a general overview of the area of application of ML in healthcare and then proceed to the specifics of symptom-based prediction, the challenges associated, and finally, the setting of healthcare booking, which drives the current work.

\section{Machine Learning in Healthcare}
\label{sec:ml_healthcare}

The use of machine learning in healthcare is over ten years old, but with the digitization of medical records and the availability of inexpensive computing power, the field has become fast-paced. \citet{Topol2019} gave an overview of the state of AI in healthcare, which showed that machine learning can be as or more effective than human clinicians in specific clinical tasks: radiological image interpretation and histopathological slide analysis.

The most common form of ML in healthcare is supervised classification: learn a model with labeled training examples (patient features + diagnosis) and apply it to give new patients a diagnosis. Naive Bayes, Logistic Regression, SVM, Decision Trees, Random Forest, and XGBoost are the most commonly used classifiers, according to which the five classifiers analyzed in this thesis can be placed: they are all supervised learning methods, but their algorithms are quite different, as they belong to different paradigms of machine learning learning approaches \citep{Sarker2021}.

According to \citet{Davenport2019}, healthcare AI was split into three domains, namely, clinical decision support, patient engagement, and administrative efficiency. The first category is disease prediction. The important thing is that these models are not supposed to substitute doctors, but to help them. It implies that they should not only be accurate but interpretable and safe too.

Among the insights that I gained during the reading of the literature is the fact that accuracy is not a sufficient measure of the performance of a medical ML model, especially when working with unbalanced medical data where it can be optimized at the expense of the overall performance of a model, that is, it can create the illusion of a model that systematically underdiagnoses a disease in a minority population. That is why I used multi-metric approach in my evaluation (Chapter~\ref{ch:results}).

\section{Symptom-Based Disease Prediction}
\label{sec:symptom_prediction}

Symptom-based disease prediction is a particular form of supervised classification in which patient-reported symptoms are the inputs and the disease category is the output. Symptom-based prediction involves less complex features than image-based or genomics-based prediction; it usually involves binary predictors of the presence or absence of each symptom. This simplicity enables it to be applied in patient-facing application and low-resource settings.

\subsection{Feature Representation Approaches}

It is common to represent most studies as binary vectors, in which each index is a symptom and the value is either 1 (present) or 0 (absent). This is easy and efficient yet it does not take into consideration severity and duration. As an illustration, a patient with a mild headache and another with a severe migraine are both represented as headache=1 and they lose the potentially useful information. Although this is a drawback, binary vectors are useful in disease classification and particularly when using ensemble methods that are able to capture the interaction of features \citep{Breiman2001}.

Other representations have been attempted by some researchers. Some studies have tested ordinal encoding in which symptoms are rated on a scale (e.g., 0 = absent, 1 = mild, 2 = moderate, 3 = severe). This will, however, involve more detailed data collection and introduces subjectivity - what one patient regards as moderate pain another may regard as severe. The resulting complexity is not necessarily correlated with the improved prediction accuracy, and that is why most research works are binary-encoded.

Symptom duration or time of onset can also be coded another way. An example of this is that a symptom beginning two days ago may be encoded differently than one beginning two weeks ago. I also failed to identify much research which has effectively utilized such type of temporal information, likely due to its difficulty in gathering it reliably and absence in most datasets.

\subsection {review of Comparative Studies.}

A number of studies have tested individual classifiers on symptom data, but the number of studies that have compared multiple approaches under the same conditions is very low. \citet{Uddin2019} compared Naive Bayes, Decision Trees, and Random Forest on multiple medical datasets and has found that ensemble methods tend to be more effective. Nevertheless, the datasets and preprocessing were different in experiments, and it is difficult to come up with solid conclusions on which classifier is indeed the best one. As an illustration, they provided the various train-test splits across the various classifiers and some of the datasets were balanced whereas others were not.

In a study by \citet{Sarker2021}, a number of classifiers were tested on a diabetes prediction dataset and it was found that the Logistic Regression was surprisingly effective compared to more advanced approaches. They explained this by the fact that the dataset has fairly easy decision boundaries which did not demand the additional modeling power of ensemble methods. This implies that the optimal classifier is dependent on the nature of the data at hand and this is one of the reasons why I test five different classifiers in this thesis and not because one is always going to win.

\citet{Chowdhury2024} carried out a systematic literature review of the ML-for-diagnosis literature and discovered that although Random Forest and XGBoost were always among the highest-ranking ones, there were very few studies that compared multiple classifiers under equal conditions. They particularly demanded controlled comparisons with the same preprocessing and evaluation procedures - which is precisely what I do in this thesis. They have reviewed 87 articles published in the years 2018-2023, but only 12 of them compared over three classifiers using the same dataset and the same preprocessing. In the vast majority of studies, a single or two classifiers were selected, optimized and good results were reported, which does not inform us whether other classifiers could have performed even better.

Previous researchers have examined the characteristics of datasets used in their work, and this is also the case for this study.

Among the things that I have observed in the course of going through the literature is that nearly all studies utilize one dataset. This leaves it impossible to determine whether the results are applicable in general or only in that data. To illustrate this, a classifier can be very good on a dataset comprising 10 disease classes and 50 symptoms and fail on a dataset comprising 40 disease classes and 132 symptoms. The number of classes and the number of features increase the problem of classification, meaning that testing on a single dataset will not provide a complete picture.

Other studies take proprietary clinical data of hospitals, and thus the results are difficult to replicate. Open datasets such as the ones found on Kaggle are simpler to access and enable other researchers to confirm them, but may be less messy or more formatted than actual clinical data. Reproducibility and realism have a trade-off.

To answer the question of generalization, I chose two datasets of extremely different sizes: one containing approximately 5,000 records \citet{Kaggle_Disease2020} and one containing approximately 246,000 \citet{Kaggle_Diseases2023}. This allows me to observe how the rankings of classifiers vary with different data scales and whether more data is useful to all of them or useful to a few more than to others.

\section{Challenges in Medical Prediction}
\label{sec:challenges}

Medical classification presents problems that cannot be encountered by standard ML benchmarks such as image classification or text classification. The three primary challenges I identified in the literature discussed in this section include class imbalance, incomplete and noisy data on symptoms, and the difficulty of evaluation.

\subsection{Class Imbalance}

The greatest obstacle is the imbalance of classes. There are common diseases (such as flu or common cold) and rare diseases (such as some autoimmune diseases), and therefore the training data is biased. The simplest approach of a classifier is choosing to predict the most frequent diseases and neglect the rare ones completely, which may result in high accuracy \citep{Haixiang2017}. To illustrate this, when 80 percent of patients in the data set have one of five common diseases, a classifier which predicts those five diseases only will achieve an 80 percent accuracy even though it is useless on the other 36 diseases.

This is the reason the accuracy itself is a misleading measure of unbalanced data. A 95 percent-accurate classifier may be performing well on all classes, or it may be cutting half of them off. That is why I employ precision, recall, and F1-score along with accuracy (see Section~\ref{sec:evaluation_metrics}).

To address class imbalance several methods have been suggested, including SMOTE (Synthetic Minority Over-sampling Technique) that was introduced by \citet{Chawla2002} and synthesizes synthetic examples of infrequent classes by interpolating between existing ones. When you have two patients of a rare disease and similar symptom patterns, SMOTE will create a new synthetic patient with a different symptom vector it between. It adds the number of training examples to rare diseases, and this enables the classifiers to learn more accurate decision boundaries.

Nonetheless, as described by \citet{Fernandez2018}, SMOTE has challenges in binary feature spaces: the interpolation-based method of generating synthetic samples using averaging of binary vectors produces fractional feature values, which do not reflect a legitimate presentation of symptoms. For instance, if patient A has symptoms [1, 0, 1, 0] and patient B has symptoms [0, 1, 1, 0], the interpolated vector might be [0.5, 0.5, 1, 0]. What does it mean for a symptom to have a value of 0.5? You either possess the symptom or not. Others represent them as 0 or 1, which may give a combination of symptoms that are not medically plausible.

Another possibility is class weighting, in which losses are more expensive to misclassify rare diseases during training. In the event that a disease occurs 100 times in the training set and the second disease occurs 10 times, you can weight the second disease 10 times that way, so that a false classification on it is equivalent to 10 false classifications of the common disease. This is supported in most scikit-learn classifiers by the \texttt{class\_weight='balanced'} parameter that automatically calculates the weights by the frequencies of the classes.

The reason why I selected class weighting in this thesis is that it operates naturally with binary features and does not produce artificial patterns of symptoms. It also does not expand the size of training set and therefore training time remains constant.

Chapter 2 of the article by Chabet (2020) experimented with various approaches to imbalance-handling (SMOTE, class weighting, undersampling, and combinations) with various classifiers and discovered that none of them are optimal in all settings. The most suitable depends on the classifier and the level of imbalance. As an example, tree-based models, such as Random Forest, are stronger against imbalance than Naive Bayes since they are able to control their splitting thresholds depending on the weights of the classes. This is among the reasons why I would go ahead and test five different classifiers as opposed to simply choosing one and believing that it would do well in all situations.

\subsection{Noisy and Incomplete Data}

In addition to imbalance, there is noisy symptom data. Patients can forget to report symptoms, report them in an imprecise way, and/or report irrelevant symptoms to their condition \citep{Chowdhury2024}. As an example, a patient with the flu may complain of headache, fever and fatigue but fail to add sore throat to the list of complaints since they do not think it is significant. This adds the missing information on the symptom vector.

Conversely, the patients may excessively report symptoms. When one searches his symptoms over the Internet and completes a questionnaire, there is a possibility that they begin to notice and report symptoms that would otherwise go unmentioned. This introduces noise in the data.

Public datasets such as those I have made use of in this thesis are less messy than actual clinical data since they have been refined and processed by the creators of the dataset. They do not usually involve the sort of sloppy, unfinished, or conflicting information you would receive by attending to actual patients. This implies that the classifier will perform optimistically on these datasets than would be the case in a real deployment.

The other problem is that symptoms may have varying meanings to different individuals. What one patient considers fatigue (I feel a bit tired) may to another be fatigue (I can hardly get out of bed). Binary symptom encoding does the same, which is information-losing. Others have attempted to overcome this by requesting patients to rate the severity of the symptoms on a scale, but this complicates the data collection process and introduces subjectivity.

\subsection{Evaluation Difficulties}

Medical prediction systems are more difficult to evaluate compared to a standard ML model. Whether you have a standard classification problem, you are concerned with the total accuracy or F1-score. Various categories of mistakes have varied implications in the field of healthcare. False negative (missing a serious disease) is significantly worse than false positive (incorrect prediction of a mild disease). False negative would cause delay in treatment whereas false positive only implies that the patient would be referred to the wrong specialist and he or she would have to be redirected.

This complicates the process of summarizing model performance in one number. You must indeed examine the confusion matrix and see what diseases are being mixed up with what. I address this further in Section~\ref{sec:error_analysis}.

The other assessment dilemma is that you cannot simply implement a system and test how it was applied into the real world, unless you are keen on ethical monitoring and regulatory acceptance. A lot of safeguards, an informed consent, and the institutional review board approval are needed to test new medical AI systems on real patients. It implies that majority of the research, such as this thesis, is based on offline testing with the help of historical data that may not be indicative of the way the system would perform in practice.

\section{Healthcare Booking Context}
\label{sec:booking_context}

The application of this work is healthcare booking. Patients are forced to choose whom to visit when they are not feeling well, whether a general practitioner, a specialist, or an emergency department. This choice is mostly made by most individuals with little medical knowledge and this results in wasted appointments, unnecessary referrals and postponements in treating serious ailments.

\subsection{The Triage Problem}

In most health systems, patients make an appointment with a general practitioner, who then proceeds to refer patients to a specialist in case of necessity. This is effective, but it is not efficient. An obviously skin-condition patient must first visit a GP and then be referred to a dermatologist, which imposes weeks of wait time on them and consumes a GP appointment that would otherwise be allocated to someone in need of general care.

Other health care providers have attempted to resolve this by using nurse-led phone lines of triage, in which a nurse asks questions about symptoms and provides advice on where to visit. This is useful, yet it still needs an individual present and this imposes limits on capacity. A prediction system based on automated symptom prediction could scale to that task at a scale, and operate 24/7, at almost zero marginal cost per patient.

The digital health strategy of the WHO also expressly favors such a tool in the context of enhancing access to healthcare, specifically in underserved regions with limited specialists available to address the needs of the population in those locations (WHO2021). When a rural clinic is able to direct patients more effectively to the appropriate type of doctor, it will decrease unnecessary traveling and waiting time.

The flow of a Booking System would be as follows.

This is how I think the practice of a symptom-based booking system would take place. A patient opens a healthcare booking application and is prompted to complete a questionnaire of symptoms. The questionnaire will list several typical symptoms (fever, cough, headache, rash, etc.) and the patient will indicate the ones he/she is experiencing.

The symptom vector is then inputted into a machine learning classifier which classifies the most probable disease type. According to that forecast, there is a recommendation of a specialist. For example:

\begin{itemize}
    \item Predicted disease: eczema → Suggested specialist: dermatologist
    \item Predicted disease: pneumonia → Suggested specialist: pulmonologist or general practitioner (urgent)
    \item Predicted disease: migraine → Suggested specialist: neurologist
\end{itemize}

The system would display the proposed specialist and the booking options. Notably it should also give the option to ignore the suggestion and make a booking with another specialist or GP in case the patient is not sure. The system is not supposed to replace decision-making, instead, it is supposed to help in the decision-making process.

\subsection {Regulatory and Ethical Concerns.}

The implementation of ML in healthcare is raising serious issues that extend beyond technical performance. The General Data Protection Regulation of the EU is very strict regarding the processing of the health data. According to GDPR, symptom information is a sensitive personal data, so the system requires explicit consent of the patient, a clear description of the use of the acquired data, and a high level of security guarantees. The patients are also entitled to view their data, revise it, and demand it to be destroyed.

Medical AI systems are considered to be of high risk by the new EU AI Act \citep{EUAI2024}, which implies that they are prone to further requirements. The system would require documented risk assessment, continuous bias or performance drift monitoring, and human control. The AI Act explicitly states that the high-risk systems must be transparent and explainable, which predisposes the models with simpler structure such as Logistic Regression in favor of the black-box models such as deep neural networks.

Liability issues also exist. In case the system recommends an incorrect specialist and a patient deteriorates due to the delay who should be held accountable? The healthcare provider? The software vendor? Who was the ML engineer that trained the model? These legal issues are yet to be sorted out in the courts and regulatory authorities.

These regulatory aspects are further elaborated in Chapter~\ref{ch:discussion}, especially their influence on the selection of classifier and system design in general.