\chapter{Discussion}
\label{ch:discussion}

The chapter explains the findings, contrasts the trade off among classifiers, admits what this paper cannot assert and the ethical and practical aspects of implementing such models.

\section{Interpretation of Results}
\label{sec:interpretation}

The experiments showed that there was no single dominating classifier in both datasets. Four of the five models performed above 98\% on all the metrics on Dataset~1, whereas on Dataset~2 the ranking shifted substantially. Logistic Regression and SVM reached perfect classification on Dataset~1. On Dataset~2, Naive Bayes achieved the best accuracy (84.11\%) and the highest macro-F1 (75.97\%), Logistic Regression and XGBoost were still competitive, and Random Forest dropped to 46.69\%. Rather than the absolute numbers, the most informative result of this comparison is the ranking shift itself.

\subsection{Dataset 1: Why All Models Scored Near-Perfect}

The near perfect scores on Dataset~1 do not imply that the classifiers have a similar performance. Only a few hundred unique records are present in Dataset~1 after deduplication, distributed over 41 disease classes, and the symptom-disease mappings in this curated set are extremely separable, with most disease classes having very few common symptoms with other classes. In that case, the test set is too simple to be able to differentiate between models, and the headline accuracy figure does not allow for any significant performance gaps to occur in a small range around 100\%. The slightly lower score of XGBoost (95.08\%) is most likely due to mild overfitting on a small training set despite L2 regularization, not for any inherent reason of the algorithm.

The implication is that the decision on which classifier to use cannot be made based on Dataset~1 alone. It is used mainly as a sanity check that the full preprocessing, encoding and evaluation pipeline is correct.

\subsection{Dataset 2: Why Naive Bayes Performed Best}

What came as a surprise is that Naive Bayes was best of all classifiers on Dataset~2, including the ensemble methods. On closer inspection, this reflects the nature of the data and the limitations of the experiment.

First, there are 727 disease classes in Dataset~2 with 377 binary features. The per-class likelihoods estimated by Bernoulli Naive Bayes are stable in this regime: for each class there are still hundreds of training examples, and the model requires only to estimate $P(x_j \mid y)$ for each feature and each class. The independence assumption acts as implicit regularization --- it helps to avoid overfitting to the higher-order interactions that are noisy and that more flexible methods attempt to fit.

Secondly, the more flexible classifiers had to run with tighter constraints on Dataset~2 than on Dataset~1. To maintain trainability at the larger scale, some hyperparameters were fixed rather than optimized (Appendix~A.2): the maximum tree depth for Random Forest was set to 10, XGBoost used 20 trees with depth 3, and SVM was approximated by a linear SGD classifier. These capacity limits are reflected in the macro-F1 gap. Logistic Regression has a high recall of 80.43\% (it benefits from class weighting) but its precision decreases to 64.50\%, which means it is better at learning minority diseases at the cost of increasing false positives. In comparison, Naive Bayes gets a better balance of precision/recall (77.18\% / 77.00\%) with none of the tuning tools.

This does not imply that Naive Bayes is always best. It means that in a high-cardinality, binary-feature, imbalanced task with limited tuning budget, a simpler probabilistic baseline can be a more powerful tool than more complex approaches. This is a discovery about the interplay of data, compute budget, and model complexity --- not a top to bottom ranking.

\subsection{Dataset 2: Why Random Forest Collapsed}

The most eye-catching drop in accuracy was seen on Dataset~2 for Random Forest, going down from 98.36\% on Dataset~1 to 46.69\%. The most likely reason is hyperparameter rather than algorithmic. Given 727 disease classes and \texttt{max\_depth=10}, one tree yields at most $2^{10} = 1024$ leaves, barely more than the number of classes. When class weighting is turned on, it is difficult to assign the necessary number of leaves to fine-grained categories and the ensemble voting performs poorly.

This gap could be closed by a deeper random forest or more estimators, if they are tuned properly. The collapse thus indicates a deployment-realistic constraint --- a limited tuning budget --- rather than a limitation of the algorithm. It also emphasizes the overall lesson of the comparison: when tuning is not available, simpler models can win by default, and when it is, models need to be tuned appropriately.

\subsection{Scaling with Data}

Reading both datasets together, the ranking of classifiers is not robust. Both Naive Bayes and Logistic Regression were competitive in both settings, while on the same fixed hyperparameters the ensemble methods ranked at the top on Dataset~1 and in the middle (XGBoost) or bottom (Random Forest) on Dataset~2. The features of the data --- e.g., the number of classes, class imbalance, and the amount of overlap in symptoms --- are as important as the algorithm.

So it is not ``the more data the merrier'' for complex models, as is typically believed. More data supports models that have a capacity matching the new problem, and that match has to be re-established by tuning, not assumed. For a deployment with the characteristics of Dataset~2 (hundreds of classes, heavy imbalance, binary features), a well-fitted Naive Bayes or Logistic Regression baseline is a reasonable initial classifier, and ensemble methods should only be adopted if empirical comparison demonstrates that they are worth the extra tuning and compute cost.

\subsection{Error Patterns}

I looked at the confusion matrices (Section~\ref{sec:error_analysis}) and found that most of the errors were made between diseases sharing similar symptoms. This is natural and not necessarily an issue for the booking use case. When the model confuses two dermatological conditions, the patient will still book a dermatologist; the error only matters when it results in the patient booking the wrong type of specialist.

There are some confused pairs that are more problematic. If the model assumes that a serious respiratory infection such as pneumonia is a milder infection such as the common cold, it may delay critical treatment. A real implementation would have to identify these high-risk confusion pairs and add safeguards; the first is to flag low-confidence predictions and direct those patients to a GP instead of a specialist.

\section{Trade-offs Between Models}
\label{sec:tradeoffs}

Raw performance data doesn't tell you the whole story. There are a number of practical considerations that factor in to the choice of classifier in a real deployment, and this section discusses the most important trade-offs.

\subsection{Interpretability}

Logistic Regression is the most transparent of the five models. The coefficients give the direct effect of the various symptoms on the prediction of the disease. For example, if the coefficient of the symptom ``yellowish\_skin'' is strongly positive for hepatitis, then the likelihood of hepatitis is raised if the person has the symptom. Logistic Regression can answer a clinician's question of why the system recommended a dermatologist, directly by displaying which symptoms made the biggest contributions to the prediction.

Random Forest and XGBoost do not offer an easy way to explain individual predictions, but do provide a feature importance ranking (Section~\ref{sec:feature_importance}). Feature importance indicates which features matter on average across all predictions, not the reasoning behind a specific prediction for a particular patient. More complex and slower-to-compute per-prediction explanations for tree-based models are available through tools like SHAP (SHapley Additive exPlanations).

Technically, Naive Bayes is an interpretable model, as it allows the inspection of class probabilities and the contribution of each symptom. However, these contributions are not always intuitive due to the independence assumption. For instance, Naive Bayes may assign a high probability to a disease even though one of its distinguishing symptoms is absent, simply because the other symptoms are common.

The EU AI Act~\citep{EUAI2024} promotes transparency of AI models. Should a clinician need to give an explanation to a regulator or auditor, Logistic Regression is much easier to explain than XGBoost, because the ``how'' or ``why'' of a particular prediction is much easier to come up with.

\subsection{Training and Prediction Speed}

Naive Bayes and Logistic Regression can be trained in seconds even on Dataset~2. This makes them suitable for scenarios in which the model is continually updated, for instance if new data comes in on a daily basis. XGBoost and Random Forest have a slower training speed, minutes to hours depending on the hyperparameter search space; their prediction times are still short enough to be used in real time.

Compared with the other methods, SVM was the slowest, because it is a one-vs-one multi-class approach that requires $C(C-1)/2$ binary classifiers for $C$ classes. With the 41 disease classes on Dataset~1, there are 820 different binary SVMs. This is not feasible for a mobile health application that needs regular updates to its model.

In a system where the model is rarely changed and trained once and left, speed isn't as important. However, when the system needs to adapt to new disease patterns, faster training is a definite plus --- for example, a new strain of flu that changes the symptom-disease relationship.

The speed of prediction is not as important for the five classifiers. Even with 200 trees, XGBoost can make a prediction in just a few milliseconds, which is quick enough for an online or mobile tool for users to interact with. Predictions of 5\,ms and 50\,ms will not be distinguishable to users.

\subsection{Handling Rare Diseases}

Random Forest with class weighting and XGBoost gave the best recall performance for the rare disease classes. Recall is the percentage of true cases that the model effectively recognizes, which is especially important for rare diseases. For a rare disease with only 50 training examples, a model that ignores the disease entirely has a minimal effect on the overall accuracy (because the disease is a very small part of the data, 1\%) but gives 0\% recall for the disease.

Naive Bayes is prone to over-predicting the common diseases on imbalanced data, since it takes class probabilities from frequencies of training data, hence rare diseases are given low prior probabilities. Although symptoms may be highly suggestive of a rare disease, Naive Bayes can predict a common one, because the prior pulls the prediction towards the more common class.

The class-weighted Logistic Regression performed better than Naive Bayes for rare diseases, but not as well as the ensemble methods. Class-weighted SVM also did reasonably on rare diseases, but is less practical because of its slow training time.

\subsection{Model Size and Deployment Constraints}

The size of the model is also important. Naive Bayes and Logistic Regression are small (a few kilobytes) and can be easily integrated into a mobile application or website. Random Forest and XGBoost models are on the larger end (several megabytes), but still manageable in today's deployment environment. Model size only becomes an issue when deploying to very resource-constrained devices.

\section{Limitations and Concerns}
\label{sec:limitations}

There are many limitations to this study that I want to address openly.

\textbf{No external validation.} This is the most significant limitation. The classifiers were tested and evaluated with cross-validation and with a test set from the same data distribution, instead of an independent clinical dataset from a real hospital or clinic. Models with good accuracy on the Kaggle data might not be as accurate when used on a different population of patients with different symptom-reporting behaviour, prevalence rates, or quality of data.

The Kaggle datasets, for instance, may have purer and more consistent symptom labels than self-reported symptoms by real patients. Real patients may have a distorted understanding of what is a symptom, forget to report symptoms, or over-report based on information they have found online. The distribution of diseases in the training data may also be different from the distribution at the deployment site --- for example, a clinic in the tropics versus a clinic in northern Europe.

Therefore, the accuracy figures given in Chapter~\ref{ch:results} are likely to be optimistic upper bounds. In a realistic deployment, I would expect performance to be several percentage points worse, until the model is re-trained on data from the real patient population.

\textbf{Simplified symptom representation.} The binary representation does not capture the severity of the symptoms, how long symptoms last, or the order in which symptoms occur. A clinician would use all of this information in diagnosis. A sudden-onset headache two hours ago is different from a mild headache that lasts for two weeks. This temporal and severity information is not available in the datasets, and this reduces the classifiers' capabilities.

There are also some symptoms that depend on the context. ``Fatigue'' can be interpreted as meaning ``I feel a little tired'' for one patient, and as ``I can hardly get out of bed'' for another. Binary encoding reduces all these to the same value and loses some of the nuance.

\textbf{Single-disease assumption.} The datasets are assumed to have a single-disease format, that is, each record contains exactly one disease label. In fact, patients often suffer from multiple diseases at the same time (comorbidities). A patient with a skin condition and diabetes would be mismanaged by a single-label classifier that has to select one or the other of these labels instead of identifying both.

Technically it is possible to do multi-label classification, but rather than predicting a single disease, the model predicts a set of diseases. This will need training data in which each patient has all of their conditions noted, not just the main one. I wasn't able to use this because my two datasets didn't support it, so I was limited to single-label classification.

\textbf{No demographic features.} Age, sex, and medical history are relevant for diagnosis but are not provided in either of the datasets. A rash in a 70-year-old can be clinically different from the same rash in a 5-year-old. Certain diseases occur more frequently in males than females or vice versa, and some previous illnesses will impact the chances of future ones.

Including demographic features would likely also enhance performance, but at the same time introduce equity and privacy issues. If the model uses age or sex to make predictions, it needs to be reviewed to make sure that it does not unfairly under-diagnose a disease --- for example, under-diagnosing a disease in women because women were under-represented in the training data for that disease.

\textbf{Hyperparameter search was not exhaustive.} I did not conduct a thorough hyperparameter search; I took the preset ranges of hyperparameters described in Appendix~\ref{app:hyperparameters}. Slightly better configurations could have been found using Bayesian optimization or larger search spaces. However, I don't think that would change the relative ranking of classifiers significantly: if XGBoost outperforms Random Forest at the chosen hyperparameters, it would probably outperform Random Forest under better-tuned ones too, since both would benefit from additional tuning.

\textbf{No cost-sensitive evaluation.} All misclassifications were given the same cost, though there may be different types of error with different costs. Predicting a mild disease incorrectly is less serious than missing a severe one. A proper deployment would assign explicit costs to various types of error, and optimize those costs instead of overall F1-score. These costs would have to be defined with input from clinicians to know which errors are most important.

\section{Ethical Considerations}
\label{sec:ethics}

Deploying disease prediction models in healthcare poses more than technical concerns. In this section, I discuss the main moral and legal issues that would have to be addressed before a real deployment.

\subsection{Patient Safety}

If a prediction is wrong, an incorrect booking will be made as well. When the model suggests a dermatologist for the patient but a cardiologist is needed, the delay could be life-threatening. In the worst case, a patient with a serious condition such as a heart attack may be referred to the wrong specialist, which means that life-saving treatment may be delayed.

When such a system is deployed, it should be obvious that the predictions it offers are not diagnoses, but rather recommendations. The interface should show something like: ``Based on your symptoms, you may want to see a dermatologist. If symptoms are severe and/or worsening, seek urgent medical attention right away.'' The system should assist in making decisions, not take the place of clinical judgment.

Another safety issue is using the system too much. Patients might become overly reliant on the AI and ignore their own intuition, or they may delay appointments while symptoms worsen. The system needs to work in the opposite way and encourage users to seek medical advice if they are unsure, rather than presenting the prediction as final.

\subsection{Data Protection and GDPR Compliance}

The GDPR~\citep{GDPR2016} classifies health data as a special category of data. Article 9 defines health information as one of the categories of sensitive personal information, which is prone to greater protection. To process health data in a symptom-based prediction system, some major provisions are required:

\begin{itemize}
    \item \textbf{Lawful basis and explicit consent (Art. 6, 9):} Patients must actively agree to symptom data processing through consent that is specific, informed, freely given, and unambiguous. Pre-checked boxes or implied consent do not meet the threshold for explicit consent.
    \item \textbf{Purpose limitation (Art. 5):} Data may only be processed for the purpose specified at collection. Symptom data collected for appointment booking cannot be repurposed for marketing analytics or secondary research without obtaining separate, specific consent.
    \item \textbf{Data minimization (Art. 5):} Systems must collect only data necessary for the stated purpose. If the classifier requires only symptom indicators, requesting unrelated personal information (e.g., postal address, non-medical contact details) violates this principle unless required for the booking function itself.
    \item \textbf{Data subject rights (Art. 15, 17):} Patients have enforceable rights to access their stored data and to request erasure under specified conditions. The system must provide technical mechanisms for fulfilling these requests within the regulatory timeframes.
    \item \textbf{Security measures (Art. 32):} Health data must be protected through appropriate technical and organizational measures, including encryption in transit (e.g., TLS/HTTPS) and at rest (e.g., encrypted storage), along with access controls limiting system interaction to authorized personnel.
\end{itemize}

An operational system would also be required to have as little data as possible stored. Ideally, no symptom data would be saved indefinitely unless the patient explicitly chooses to preserve their history. The anonymization of logs and analytics must be done so that individual patients cannot be re-identified.

\subsection{EU AI Act Requirements}

The EU AI Act~\citep{EUAI2024} lists medical diagnostic AI systems in Annex III (category 5b: AI systems intended to assist in medical diagnosis or treatment) as high-risk systems, because of the potentially serious consequences resulting from their use. This classification triggers a number of mandatory compliance requirements for any commercial deployment:

\begin{itemize}
    \item \textbf{Conformity assessment (Art. 43):} Before being placed on the market, high-risk AI systems must undergo a conformity assessment confirming that they meet safety and performance requirements. This requires detailed technical documentation covering system design, training data, validation procedures, and risk mitigation measures.
    \item \textbf{Risk management system (Art. 9):} Providers must define and operate documented risk-management processes across the system lifecycle. This includes identifying reasonably foreseeable risks (incorrect predictions, algorithmic bias, security vulnerabilities), mitigating them, and establishing post-market monitoring to detect performance degradation.
    \item \textbf{Human oversight (Art. 14):} High-risk systems must be designed to allow effective human oversight, through human-in-the-loop or human-on-the-loop mechanisms. For a symptom-based booking system, this could be implemented as clinical review of low-confidence predictions or as periodic audits of specialist referral patterns.
    \item \textbf{Transparency obligations (Art. 13):} Systems must provide users with clear, accessible information about system capabilities, operating logic, data usage, and known limitations. These requirements favour interpretable architectures (e.g., Logistic Regression) over opaque models that offer only limited explanations.
    \item \textbf{Accuracy and robustness (Art. 15):} Systems must reach and maintain accuracy levels appropriate for their intended purpose. Providers are responsible for continuous performance monitoring and model maintenance, including retraining when accuracy metrics degrade.
\end{itemize}

These are legally binding requirements. Non-compliance with the EU AI Act is punishable by administrative fines of up to 15 million Euro or 3\% of total worldwide annual turnover, whichever is greater.

The EU AI Act also requires technical documentation of the training data, model architecture, hyperparameters, and performance metrics. This implies that all aspects of the methodology described in this thesis (Chapter~\ref{ch:methodology} and Appendix~\ref{app:supplementary}) would need to be maintained in a real deployment, not solely for academic use.

\subsection{Algorithmic Bias and Fairness}

Sometimes the algorithm may be unfair and biased. If the training data is unrepresentative of certain groups, then the model will be less effective for those groups. To illustrate, if the dataset consists predominantly of young adults, the model may not be as effective in elderly patients, whose symptoms may differ due to age-related comorbidities.

I used two datasets that both lack demographic information (age, sex and ethnicity), so I cannot evaluate if the classifiers are biased along these dimensions. It is important to note that a fairness audit of accuracy, precision and recall by age, sex, and ethnicity would be needed before deployment of such a system, to verify that performance is similar among population groups.

In the event that the model was found to underperform on certain groups, it would infringe on the non-discrimination clauses of the GDPR and the fairness clauses of the EU AI Act. The provider would then have to gather more representative training data or adjust the model (for example, by using fairness-aware training methods) to level performance between groups.

Another type of bias is language and health literacy. If the symptom questionnaire is written in medical jargon, less health-literate patients might not be able to interpret the questions or respond adequately, which would give poor predictions for that population. The interface design is therefore as crucial as the model design.

\section{Practical Deployment Considerations}
\label{sec:deployment}

The results of the experiments presented here are not the only steps required to convert the system to an operational configuration; the remaining steps are not addressed in this thesis but are outlined here. This section sketches what an actual implementation would look like and the issues that it would have to consider.

\subsection{System Architecture}

The system should be made up of several components:

\begin{enumerate}
    \item \textbf{User interface:} A web or mobile application that presents patients with a symptom checklist. The interface should be simple and use plain language rather than medical jargon. The patient checks a box for each symptom they are experiencing.
    \item \textbf{Feature encoding:} The selected symptoms are encoded as a binary vector matching the format used during training (132 features, one per symptom).
    \item \textbf{Classifier:} The trained XGBoost or Random Forest model is applied to the feature vector and produces a predicted disease together with a confidence score (probability).
    \item \textbf{Disease-to-specialist mapping:} A lookup table maps each disease to the appropriate specialist (for example, eczema $\rightarrow$ dermatologist, pneumonia $\rightarrow$ pulmonologist, migraine $\rightarrow$ neurologist). This mapping is defined by clinicians rather than learned by the model.
    \item \textbf{Booking system integration:} The system queries the available appointment slots for the recommended specialist and presents them to the patient.
\end{enumerate}

The entire process, from the symptom checklist through to the specialist recommendation, should take less than a second to maintain the user experience.

\subsection{Confidence Thresholds}

Not all predictions are necessarily to be followed. If the model is in doubt, the patient should be referred to a GP (General Practitioner) and not straight to a specialist. Making such a threshold is a compromise between referral that is not required and missed diagnosis, which is a decision that should be made by clinicians, not only engineers.

For instance, if the model says it can predict the label ``diabetes'' with 85\% confidence, then it is reasonable to recommend that the patient be referred to an endocrinologist. However, if the model assigns only 55\% likelihood to diabetes, the symptoms are ambiguous, and the system should encourage the user to seek a GP appointment to be properly evaluated, instead of risking a potentially erroneous specialist referral.

This can be achieved by establishing a fixed level of confidence (such as 70\%) below which all patients are referred to a GP. It would be necessary to determine the exact threshold via pilot testing on actual patients and feedback from clinicians on the rate of inappropriate referrals.

Another approach is to take advantage of the uncertainty of the model: if the top two predicted diseases are extremely close (e.g., 52\% vs. 48\%), the uncertainty is high and the system should be cautious. If the top prediction is 85\% and the second prediction is only 10\%, the signal is more powerful and the system can suggest the specialist with greater confidence.

\subsection{Handling Edge Cases}

There would be a number of edge cases to cover:

\begin{itemize}
    \item \textbf{Emergency symptoms:} If the patient reports symptoms that may indicate a medical emergency (e.g., chest pain, difficulty breathing, severe bleeding), the system must override the classifier and direct the patient to emergency services or an emergency department immediately. This can be implemented with simple rule-based logic applied before the ML model.
    \item \textbf{No symptoms selected:} If the patient does not check any symptoms, the system cannot make a prediction. It should require the user to select at least one symptom, or provide a free-text entry field to describe their condition.
    \item \textbf{Very rare symptom combinations:} If the patient reports a combination of symptoms that does not appear in the training data, the model's prediction may be unreliable. The system could detect this by computing the distance to the nearest training example and flag cases that lie far from anything the model has seen before.
\end{itemize}

\subsection{Monitoring and Maintenance}

A deployed model would require ongoing monitoring. Disease patterns shift, new diseases emerge (as with COVID-19), and the patient population using the system may not match the original training data. Periodic retraining and performance reviews are needed to detect degradation before it can harm patients.

Several metrics should be monitored:

\begin{itemize}
    \item \textbf{Prediction accuracy:} A rising rate of incorrect predictions signals a problem. This requires feedback from clinicians on whether the specialist referrals were appropriate.
    \item \textbf{Prediction distribution:} If the model suddenly starts predicting one disease much more often than usual, this may indicate either a real disease outbreak or a data shift. For example, an increase in flu predictions during flu season is expected; an increase in predictions of rare diseases is more likely a data-quality problem.
    \item \textbf{User feedback:} The system should ask users whether the specialist recommendation was useful. A pattern of users reporting unhelpful suggestions is an indicator that requires investigation.
    \item \textbf{Coverage:} What fraction of users receive a high-confidence prediction versus being routed to a GP? If too many users are routed to a GP because of low-confidence predictions, the model is no longer providing value.
\end{itemize}

Retraining should be performed periodically --- for example, quarterly --- or whenever performance drops below a chosen threshold. Retraining involves collecting new data (anonymized patient records with symptoms and confirmed diagnoses), adding it to the training set, and re-running the hyperparameter search and evaluation pipeline reported in this thesis.

\subsection{Integration with Existing Systems}

A real implementation would have to interoperate with existing healthcare IT systems. This includes:

\begin{itemize}
    \item \textbf{Electronic health records (EHR):} If the patient's medical history is available, it can be supplied as additional features to improve predictions. For example, a patient with a history of heart disease should be routed differently from a patient with no relevant medical history.
    \item \textbf{Appointment scheduling systems:} The system must be able to query available appointment slots and allow patients to book directly.
    \item \textbf{Insurance and billing:} The system may also need to verify whether visits to the recommended specialist are covered by the patient's insurance.
\end{itemize}

These integrations are not technically complex, but they require coordination with multiple stakeholders and adherence to healthcare IT standards such as HL7 or FHIR.